---
title: "Class 2 Age Guess Analysis"
author: "Thomas E. Love"
date: "August 31, 2017"
output: 
    html_document:
        code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

```{r load-packages}
library(magrittr); library(tidyverse)
```

```{r load-dataset}
age_love_2017 <- read_csv("age-love-2017.csv")
```

```{r look-at-data}
age_love_2017
```

Our data set has `r nrow(age_love_2017)` rows, and `r ncol(age_love_2017)` columns.

# Build a Picture of the First Guesses

```{r first-histogram}
ggplot(data = age_love_2017, aes(x = age1)) +
    geom_histogram()
```

## A nicer picture

```{r second-histogram}
ggplot(data = age_love_2017, aes(x = age1)) +
    geom_histogram(bins = 12, fill = "royalblue", col = "green") +
    labs(x = "First Guess of Dr. Love's Age on 2017-08-29",
         y = "# of students",
         title = "2017 First Guesses of Dr. Love's Age",
         subtitle = "His true age was 50.5")
```

# A Numerical Summary

```{r numerical-summary}
summary(age_love_2017)
```

# Better First or Second Guess?

```{r calculate-errors}
age_love_2017 <- age_love_2017 %>%
    mutate(error1 = abs(age1 - 50.5),
           error2 = abs(age2 - 50.5))

summary(age_love_2017)
```

# Compare the Guesses

```{r guess-1-vs-2}
ggplot(data = age_love_2017, aes(x = error1, y = error2)) +
    geom_point()
```

## Add a Prediction Model?

```{r guess1-vs-guess2-with-loess-smooth}
ggplot(data = age_love_2017, aes(x = age1, y = age2)) +
    geom_point(size = 3) + 
    geom_smooth(method = "loess")
```

# Plot the (matched) differences

```{r histogram-of-differences}
ggplot(age_love_2017, aes(x = age1 - age2)) +
    geom_histogram(binwidth = 2, 
                   col = "green",
                   fill = "royalblue")
```

# Numerical Summary of the Difference in Ages

```{r comparing-guess1-to-guess2}
age_love_2017 %$%
    summary(age1 - age2)
```

# How many people thought I looked younger in guess 2?

```{r count-of-younger-guesses}
age_love_2017 %>%
    count(age1-age2 < 0)
```

I think this is good news! Maybe ...

# The Much-Dreaded t test

```{r t-test-comparing-guess1-to-guess2}
age_love_2017 %$%
    t.test(age1 - age2)
```

