---
title: "Demonstration Project"
author: "TEL & Rujia Liu"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
    number_sections: TRUE
    df_print: paged
---

```{r packages, message=FALSE, warning=FALSE}
library(knitr); library(lme4); library(broom)
library(magrittr); library(tidyverse)

## Global options
opts_chunk$set(comment=NA)
opts_knit$set(width=80)
```

```{r setup, include=FALSE}
decim <- function(x, k) format(round(x, k), nsmall=k)
thou <- function(x) format(round(x, 0), big.mark = ",")
```

# Data Setup

The `hbpdat` file contains the most recent master data.

```{r data_ingest, warning=FALSE}
hbpraw <- read.csv("master htn 2017-12-05.csv") %>% tbl_df

hbpdat <- hbpraw %>%
  filter(CATVAR %in% c("Overall", "INSUR", "RACE", "GENDER",
                       "INCOMECAT", "HSGRADCAT", "LANG", "LOCATION"))

hbpdat$time <- 2006.5 + (hbpdat$REPORT/2)
hbpdat$years <- hbpdat$time - 2012.5 # years since 2012-13 report
hbpdat$incomeK <- hbpdat$INC_MEAN/1000

hbpdat$category <- fct_recode(hbpdat$CATVAR,
                              "Insurance" = "INSUR",
                             "Place of Residence" = "LOCATION",
                             "Education" = "HSGRADCAT",
                             "Sex" = "GENDER",
                             "Race-Ethnicity" = "RACE",
                             "Language" = "LANG",
                             "Income" = "INCOMECAT")

hbpdat$subgroup <- fct_recode(hbpdat$CATTXT,
     "African-Am." = "african-aO",
     "Cleveland" = "CLEVE",
     "Commercial" = "COMMERCIAL",
     "Cuyahoga Suburbs" = "CUYA NOT CLEVE",
     "Prefer English" = "ENGLISH",
     "Female" = "FEMALE",
     "High Education" = "HIGH EDUC",
     "High Income" = "HIGH INCOME",
     "Hispanic/Latino" = "hispanicO",
     "Low Education" = "LOW EDUC",
     "Low Income" = "LOW INCOME",
     "Male" = "MALE",
     "Medicaid" = "MEDICAID",
     "Medicare" = "MEDICARE",
     "Middle Education" = "MIDDLE EDUC",
     "Middle Income" = "MIDDLE INCOME",
     "Outside Cuyahoga" = "OUTSIDE CUYA",
     "ALL PATIENTS" = "OVERALL",
     "Prefer Spanish" = "SPANISH",
     "Uninsured" = "UNINSURED",
     "White" = "whiteO")
```

## Analytic Decisions Implemented in this Document

These issues will need to be discussed on 2017-12-13.

1. Include data from 2012-13 through 2016-17.
2. Include data from all practices at all systems.
3. Include data from a practice if it has at least 50 HBP subjects.
4. Include data from a practice in subgroup analyses only if it has at least 50 HBP subjects in that subgroup.
5. Covariate adjustment is time-varying, rather than baseline only.
6. Clustering by practice, but no adjustment for practice's baseline level of BP < 140/90.
7. Time is measured in calendar time, so this is not a "Time 0" analysis.
8. Fit linear, rather than piecewise linear or quadratic or spline models.

# Mixed Effects Models for BP Control (Analysis 1)

The goal here is to describe overall change in BP < 140/90 from 2012-13 through 2016-17 across the region, first in an unadjusted analysis, then in a covariate-adjusted analysis. We could present separate estimates overall and within insurance groups.

## Across All Patients (Overall subgroup)

### Select out the overall practice data from 2012-13 and forward, only those practices with at least 50 patients.

```{r}
hbp_overall <- hbpdat %>%
  filter(PRACTICE != "System") %>%
  filter(PRACTICE != "Region") %>%
  filter(SYSTEM != "SNET") %>%
  filter(time >= 2012.5) %>%
  filter(HBPPTS > 49) %>%
  filter(ORDER == 100) %>%
  select(REPORT, time, years, SYSTEM, PRACTICE, PRACCODE, ORDER, category, subgroup,
         HBPPTS, BPLT14090, BMILT30, AGE_MEAN, FEMALE, incomeK, ED_MEAN, WHITE, 
         MEDICARE, COMMERCIAL, MEDICAID, UNINSURED, ENGLISH, SPANISH, 
         CLEVE, CUYNOTCLE, NOTCUYA ) %>% 
  droplevels()
```

### Descriptive Summaries (counting only practices with at least 50 HBP patients)

```{r}
hbp_overall <- hbp_overall %>%
  mutate(n_BPLT14090 = round(HBPPTS * BPLT14090 / 100, 0))

tab_d_overall <- hbp_overall %>% group_by(time, years) %>%
  summarize(Practices = n(), Patients = sum(HBPPTS), 
            Pct_below_14090 = 100*sum(n_BPLT14090)/sum(HBPPTS))

kable(tab_d_overall)
```

### Figure 1. Achievement of BP < 140/90, Region-wide, All Patients

```{r}
ggplot(data = hbp_overall, aes(x = time, y = BPLT14090, size = HBPPTS)) +
  geom_point(pch = 1) +
  guides(size = FALSE) +
  geom_line(data = tab_d_overall, aes(x = time, y = Pct_below_14090), 
            col = "blue", size = 1) +
  geom_label(data = tab_d_overall, aes(x = time, y = Pct_below_14090, 
                                  label = decim(Pct_below_14090,1)), 
             col = "blue", size = 3) +
  theme_bw() + 
  labs(x = "Better Health Partnership Reporting Period",
       y = "% of Patients with Blood Pressure below 140/90",
       title = "Blood Pressure below < 140/90: July 2012 - June 2017, All Patients",
       subtitle = "Regional (blue) and Practice Achievement",
       caption = "Practice size proportional to circle size, region in blue, minimum 50 patients per practice")

```

### Model 1. No covariates model, assuming linear time effect

This model predicts the percentage of patients with BP < 140/90 at a particular moment in time, using time (in years) since 2012-13, accounting for clustering by practice, and weighting by the number of patients. It's a weighted linear mixed model.

```{r}
mod1_overall <- hbp_overall %$%
  lmer(BPLT14090 ~ years + (1 | PRACCODE), weights = HBPPTS)

# summary(mod1_overall)
# confint(mod1_overall, method = "Wald")
```

### Model 2. Time-varying covariate adjusted model, linear time

This model predicts the percentage of patients with BP < 140/90 at a particular moment in time, using time (in years) since 2012-13, accounting for clustering by practice, and weighting by the number of patients, and using as covariates the following values that vary over time:

- % of patients with BMI < 30
- mean age of patients in the practice
- % of patients in the practice who are female
- neighborhood median income estimate (in $1000s)
- neighborhood high school graduation rate estimate
- % of patients who are white
- % of patients who have commercial insurance
- % of patients who have Medicaid insurance
- % of patients who are uninsured
- % of patients who prefer English
- % of patients who live in the Cuyahoga County suburbs
- % of patients who live outside of Cuyahoga County

It, too is a weighted linear mixed model.

```{r}
mod2_overall <- hbp_overall %$%
  lmer(BPLT14090 ~ years + BMILT30 + AGE_MEAN + 
         FEMALE + incomeK + ED_MEAN + WHITE + 
         COMMERCIAL + MEDICAID + UNINSURED + 
         ENGLISH + CUYNOTCLE + NOTCUYA + 
         (1 | PRACCODE), weights = HBPPTS)

# summary(mod2_overall)
# confint(mod2_overall, method = "Wald")
```

## Models 1 and 2 restricted to Medicare patients

### Select out the Medicare practice data from 2012-13 and forward, only those practices with at least 50 Medicare patients.

```{r}
hbp_medicare <- hbpdat %>%
  filter(PRACTICE != "System") %>%
  filter(PRACTICE != "Region") %>%
  filter(SYSTEM != "SNET") %>%
  filter(time >= 2012.5) %>%
  filter(HBPPTS > 49) %>%
  filter(ORDER == 101) %>%
  select(REPORT, time, years, SYSTEM, PRACTICE, PRACCODE, ORDER, category, subgroup,
         HBPPTS, BPLT14090, BMILT30, AGE_MEAN, FEMALE, incomeK, ED_MEAN, WHITE, 
         MEDICARE, COMMERCIAL, MEDICAID, UNINSURED, ENGLISH, SPANISH, 
         CLEVE, CUYNOTCLE, NOTCUYA ) %>% 
  droplevels()
```

### Descriptive Summaries (counting only practices with at least 50 Medicare HBP patients)

```{r}
hbp_medicare <- hbp_medicare %>%
  mutate(n_BPLT14090 = round(HBPPTS * BPLT14090 / 100, 0))

tab_d_medicare <- hbp_medicare %>% group_by(time, years) %>%
  summarize(Practices = n(), Patients = sum(HBPPTS), 
            Pct_below_14090 = 100*sum(n_BPLT14090)/sum(HBPPTS))

kable(tab_d_medicare)
```

### Figure 2. Achievement of BP < 140/90, Region-wide, Medicare Patients

```{r}
ggplot(data = hbp_medicare, aes(x = time, y = BPLT14090, size = HBPPTS)) +
  geom_point(pch = 1) +
  guides(size = FALSE) +
  geom_line(data = tab_d_medicare, aes(x = time, y = Pct_below_14090), 
            col = "blue", size = 1) +
  geom_label(data = tab_d_medicare, aes(x = time, y = Pct_below_14090, 
                                  label = decim(Pct_below_14090,1)), 
             col = "blue", size = 3) +
  theme_bw() + 
  labs(x = "Better Health Partnership Reporting Period",
       y = "% of Patients with Blood Pressure below 140/90",
       title = "Blood Pressure below < 140/90: July 2012 - June 2017, Medicare Patients",
       subtitle = "Regional (blue) and Practice Achievement",
       caption = "Practice size proportional to circle size, region in blue, minimum 50 Medicare patients per practice")

```

### Model 1. No covariates model, assuming linear time effect

```{r}
mod1_medicare <- hbp_medicare %$%
  lmer(BPLT14090 ~ years + (1 | PRACCODE), weights = HBPPTS)

# summary(mod1_medicare)
# confint(mod1_medicare, method = "Wald")
```

### Model 2. Time-varying covariate adjusted model, linear time

This model predicts the percentage of patients with BP < 140/90 at a particular moment in time, using time (in years) since 2012-13, accounting for clustering by practice, and weighting by the number of patients, and using as covariates the following values that vary over time:

- % of patients with BMI < 30
- mean age of patients in the practice
- % of patients in the practice who are female
- neighborhood median income estimate (in $1000s)
- neighborhood high school graduation rate estimate
- % of patients who are white
- % of patients who prefer English
- % of patients who live in the Cuyahoga County suburbs
- % of patients who live outside of Cuyahoga County

BUT EXCLUDING the insurance information.

It, too is a weighted linear mixed model.

```{r}
mod2_medicare <- hbp_medicare %$%
  lmer(BPLT14090 ~ years + BMILT30 + AGE_MEAN + 
         FEMALE + incomeK + ED_MEAN + WHITE + 
         ENGLISH + CUYNOTCLE + NOTCUYA + 
         (1 | PRACCODE), weights = HBPPTS)

# summary(mod2_medicare)
# confint(mod2_medicare, method = "Wald")
```

## Models 1 and 2 restricted to Commercial patients

### Select out the Commercial practice data from 2012-13 and forward, only those practices with at least 50 Commercial patients.

```{r}
hbp_commercial <- hbpdat %>%
  filter(PRACTICE != "System") %>%
  filter(PRACTICE != "Region") %>%
  filter(SYSTEM != "SNET") %>%
  filter(time >= 2012.5) %>%
  filter(HBPPTS > 49) %>%
  filter(ORDER == 102) %>%
  select(REPORT, time, years, SYSTEM, PRACTICE, PRACCODE, ORDER, category, subgroup,
         HBPPTS, BPLT14090, BMILT30, AGE_MEAN, FEMALE, incomeK, ED_MEAN, WHITE, 
         MEDICARE, COMMERCIAL, MEDICAID, UNINSURED, ENGLISH, SPANISH, 
         CLEVE, CUYNOTCLE, NOTCUYA ) %>% 
  droplevels()
```

### Descriptive Summaries (counting only practices with at least 50 Commercial HBP patients)

```{r}
hbp_commercial <- hbp_commercial %>%
  mutate(n_BPLT14090 = round(HBPPTS * BPLT14090 / 100, 0))

tab_d_commercial <- hbp_commercial %>% group_by(time, years) %>%
  summarize(Practices = n(), Patients = sum(HBPPTS), 
            Pct_below_14090 = 100*sum(n_BPLT14090)/sum(HBPPTS))

kable(tab_d_commercial)
```

### Figure 3. Achievement of BP < 140/90, Region-wide, Commercial Patients

```{r}
ggplot(data = hbp_commercial, aes(x = time, y = BPLT14090, size = HBPPTS)) +
  geom_point(pch = 1) +
  guides(size = FALSE) +
  geom_line(data = tab_d_commercial, aes(x = time, y = Pct_below_14090), 
            col = "blue", size = 1) +
  geom_label(data = tab_d_commercial, aes(x = time, y = Pct_below_14090, 
                                  label = decim(Pct_below_14090,1)), 
             col = "blue", size = 3) +
  theme_bw() + 
  labs(x = "Better Health Partnership Reporting Period",
       y = "% of Patients with Blood Pressure below 140/90",
       title = "Blood Pressure below < 140/90: July 2012 - June 2017, Commercial Patients",
       subtitle = "Regional (blue) and Practice Achievement",
       caption = "Practice size proportional to circle size, region in blue, minimum 50 Commercial patients per practice")

```

### Model 1. No covariates model, assuming linear time effect

```{r}
mod1_commercial <- hbp_commercial %$%
  lmer(BPLT14090 ~ years + (1 | PRACCODE), weights = HBPPTS)

# summary(mod1_commercial)
# confint(mod1_commercial, method = "Wald")
```

### Model 2. Time-varying covariate adjusted model, linear time

This model predicts the percentage of patients with BP < 140/90 at a particular moment in time, using time (in years) since 2012-13, accounting for clustering by practice, and weighting by the number of patients, and using as covariates the following values that vary over time:

- % of patients with BMI < 30
- mean age of patients in the practice
- % of patients in the practice who are female
- neighborhood median income estimate (in $1000s)
- neighborhood high school graduation rate estimate
- % of patients who are white
- % of patients who prefer English
- % of patients who live in the Cuyahoga County suburbs
- % of patients who live outside of Cuyahoga County

BUT EXCLUDING the insurance information.

It, too is a weighted linear mixed model.

```{r}
mod2_commercial <- hbp_commercial %$%
  lmer(BPLT14090 ~ years + BMILT30 + AGE_MEAN + 
         FEMALE + incomeK + ED_MEAN + WHITE + 
         ENGLISH + CUYNOTCLE + NOTCUYA + 
         (1 | PRACCODE), weights = HBPPTS)

# summary(mod2_commercial)
# confint(mod2_commercial, method = "Wald")
```

## Models 1 and 2 restricted to Medicaid patients

### Select out the Medicaid practice data from 2012-13 and forward, only those practices with at least 50 Medicaid patients.

```{r}
hbp_medicaid <- hbpdat %>%
  filter(PRACTICE != "System") %>%
  filter(PRACTICE != "Region") %>%
  filter(SYSTEM != "SNET") %>%
  filter(time >= 2012.5) %>%
  filter(HBPPTS > 49) %>%
  filter(ORDER == 103) %>%
  select(REPORT, time, years, SYSTEM, PRACTICE, PRACCODE, ORDER, category, subgroup,
         HBPPTS, BPLT14090, BMILT30, AGE_MEAN, FEMALE, incomeK, ED_MEAN, WHITE, 
         MEDICARE, COMMERCIAL, MEDICAID, UNINSURED, ENGLISH, SPANISH, 
         CLEVE, CUYNOTCLE, NOTCUYA ) %>% 
  droplevels()
```

### Descriptive Summaries (counting only practices with at least 50 Medicaid HBP patients)

```{r}
hbp_medicaid <- hbp_medicaid %>%
  mutate(n_BPLT14090 = round(HBPPTS * BPLT14090 / 100, 0))

tab_d_medicaid <- hbp_medicaid %>% group_by(time, years) %>%
  summarize(Practices = n(), Patients = sum(HBPPTS), 
            Pct_below_14090 = 100*sum(n_BPLT14090)/sum(HBPPTS))

kable(tab_d_medicaid)
```

### Figure 4. Achievement of BP < 140/90, Region-wide, Medicaid Patients

```{r}
ggplot(data = hbp_medicaid, aes(x = time, y = BPLT14090, size = HBPPTS)) +
  geom_point(pch = 1) +
  guides(size = FALSE) +
  geom_line(data = tab_d_medicaid, aes(x = time, y = Pct_below_14090), 
            col = "blue", size = 1) +
  geom_label(data = tab_d_medicaid, aes(x = time, y = Pct_below_14090, 
                                  label = decim(Pct_below_14090,1)), 
             col = "blue", size = 3) +
  theme_bw() + 
  labs(x = "Better Health Partnership Reporting Period",
       y = "% of Patients with Blood Pressure below 140/90",
       title = "Blood Pressure below < 140/90: July 2012 - June 2017, Medicaid Patients",
       subtitle = "Regional (blue) and Practice Achievement",
       caption = "Practice size proportional to circle size, region in blue, minimum 50 Medicaid patients per practice")

```

### Model 1. No covariates model, assuming linear time effect

```{r}
mod1_medicaid <- hbp_medicaid %$%
  lmer(BPLT14090 ~ years + (1 | PRACCODE), weights = HBPPTS)

# summary(mod1_medicaid)
# confint(mod1_medicaid, method = "Wald")
```

### Model 2. Time-varying covariate adjusted model, linear time

This model predicts the percentage of patients with BP < 140/90 at a particular moment in time, using time (in years) since 2012-13, accounting for clustering by practice, and weighting by the number of patients, and using as covariates the following values that vary over time:

- % of patients with BMI < 30
- mean age of patients in the practice
- % of patients in the practice who are female
- neighborhood median income estimate (in $1000s)
- neighborhood high school graduation rate estimate
- % of patients who are white
- % of patients who prefer English
- % of patients who live in the Cuyahoga County suburbs
- % of patients who live outside of Cuyahoga County

BUT EXCLUDING the insurance information.

It, too is a weighted linear mixed model.

```{r}
mod2_medicaid <- hbp_medicaid %$%
  lmer(BPLT14090 ~ years + BMILT30 + AGE_MEAN + 
         FEMALE + incomeK + ED_MEAN + WHITE + 
         ENGLISH + CUYNOTCLE + NOTCUYA + 
         (1 | PRACCODE), weights = HBPPTS)

# summary(mod2_medicaid)
# confint(mod2_medicaid, method = "Wald")
```


## Models 1 and 2 restricted to Uninsured patients

### Select out the Uninsured practice data from 2012-13 and forward, only those practices with at least 50 Uninsured patients.

```{r}
hbp_uninsured <- hbpdat %>%
  filter(PRACTICE != "System") %>%
  filter(PRACTICE != "Region") %>%
  filter(SYSTEM != "SNET") %>%
  filter(time >= 2012.5) %>%
  filter(HBPPTS > 49) %>%
  filter(ORDER == 104) %>%
  select(REPORT, time, years, SYSTEM, PRACTICE, PRACCODE, ORDER, category, subgroup,
         HBPPTS, BPLT14090, BMILT30, AGE_MEAN, FEMALE, incomeK, ED_MEAN, WHITE, 
         MEDICARE, COMMERCIAL, MEDICAID, UNINSURED, ENGLISH, SPANISH, 
         CLEVE, CUYNOTCLE, NOTCUYA ) %>% 
  droplevels()
```

### Descriptive Summaries (counting only practices with at least 50 Uninsured HBP patients)

```{r}
hbp_uninsured <- hbp_uninsured %>%
  mutate(n_BPLT14090 = round(HBPPTS * BPLT14090 / 100, 0))

tab_d_uninsured <- hbp_uninsured %>% group_by(time, years) %>%
  summarize(Practices = n(), Patients = sum(HBPPTS), 
            Pct_below_14090 = 100*sum(n_BPLT14090)/sum(HBPPTS))

kable(tab_d_uninsured)
```

### Figure 4. Achievement of BP < 140/90, Region-wide, Uninsured Patients

```{r}
ggplot(data = hbp_uninsured, aes(x = time, y = BPLT14090, size = HBPPTS)) +
  geom_point(pch = 1) +
  guides(size = FALSE) +
  geom_line(data = tab_d_uninsured, aes(x = time, y = Pct_below_14090), 
            col = "blue", size = 1) +
  geom_label(data = tab_d_uninsured, aes(x = time, y = Pct_below_14090, 
                                  label = decim(Pct_below_14090,1)), 
             col = "blue", size = 3) +
  theme_bw() + 
  labs(x = "Better Health Partnership Reporting Period",
       y = "% of Patients with Blood Pressure below 140/90",
       title = "Blood Pressure below < 140/90: July 2012 - June 2017, Uninsured Patients",
       subtitle = "Regional (blue) and Practice Achievement",
       caption = "Practice size proportional to circle size, region in blue, minimum 50 Uninsured patients per practice")

```

### Model 1. No covariates model, assuming linear time effect

```{r}
mod1_uninsured <- hbp_uninsured %$%
  lmer(BPLT14090 ~ years + (1 | PRACCODE), weights = HBPPTS)

# summary(mod1_uninsured)
# confint(mod1_uninsured, method = "Wald")
```

### Model 2. Time-varying covariate adjusted model, linear time

This model predicts the percentage of patients with BP < 140/90 at a particular moment in time, using time (in years) since 2012-13, accounting for clustering by practice, and weighting by the number of patients, and using as covariates the following values that vary over time:

- % of patients with BMI < 30
- mean age of patients in the practice
- % of patients in the practice who are female
- neighborhood median income estimate (in $1000s)
- neighborhood high school graduation rate estimate
- % of patients who are white
- % of patients who prefer English
- % of patients who live in the Cuyahoga County suburbs
- % of patients who live outside of Cuyahoga County

BUT EXCLUDING the insurance information.

It, too is a weighted linear mixed model.

```{r}
mod2_uninsured <- hbp_uninsured %$%
  lmer(BPLT14090 ~ years + BMILT30 + AGE_MEAN + 
         FEMALE + incomeK + ED_MEAN + WHITE + 
         ENGLISH + CUYNOTCLE + NOTCUYA + 
         (1 | PRACCODE), weights = HBPPTS)

# summary(mod2_uninsured)
# confint(mod2_uninsured, method = "Wald")
```


## Results for Models 1 and 2

### Build the data set of results

```{r}
res12 <- data.frame(
  model = rep(c(1, 2)),
  subgroup = c(rep("All Patients", 2), rep("Medicare", 2), rep("Commercial", 2),
               rep("Medicaid",2), rep("Uninsured",2)),
  estimate = c(filter(tidy(mod1_overall), term == "years")$estimate,
               filter(tidy(mod2_overall), term == "years")$estimate,
               filter(tidy(mod1_medicare), term == "years")$estimate,
               filter(tidy(mod2_medicare), term == "years")$estimate,
               filter(tidy(mod1_commercial), term == "years")$estimate,
               filter(tidy(mod2_commercial), term == "years")$estimate,
               filter(tidy(mod1_medicaid), term == "years")$estimate,
               filter(tidy(mod2_medicaid), term == "years")$estimate,
               filter(tidy(mod1_uninsured), term == "years")$estimate,
               filter(tidy(mod2_uninsured), term == "years")$estimate),
  std.err = c(filter(tidy(mod1_overall), term == "years")$std.err,
              filter(tidy(mod2_overall), term == "years")$std.err,
              filter(tidy(mod1_medicare), term == "years")$std.err,
              filter(tidy(mod2_medicare), term == "years")$std.err,
              filter(tidy(mod1_commercial), term == "years")$std.err,
              filter(tidy(mod2_commercial), term == "years")$std.err,
              filter(tidy(mod1_medicaid), term == "years")$std.err,
              filter(tidy(mod2_medicaid), term == "years")$std.err,
              filter(tidy(mod1_uninsured), term == "years")$std.err,
              filter(tidy(mod2_uninsured), term == "years")$std.err),
  low.95ci = c(confint(mod1_overall, method = "Wald")["years",1],
               confint(mod2_overall, method = "Wald")["years",1],
               confint(mod1_medicare, method = "Wald")["years",1],
               confint(mod2_medicare, method = "Wald")["years",1],
               confint(mod1_commercial, method = "Wald")["years",1],
               confint(mod2_commercial, method = "Wald")["years",1],
               confint(mod1_medicaid, method = "Wald")["years",1],
               confint(mod2_medicaid, method = "Wald")["years",1],
               confint(mod1_uninsured, method = "Wald")["years",1],
               confint(mod2_uninsured, method = "Wald")["years",1]),
    hi.95ci = c(confint(mod1_overall, method = "Wald")["years",2],
              confint(mod2_overall, method = "Wald")["years",2],
              confint(mod1_medicare, method = "Wald")["years",2],
              confint(mod2_medicare, method = "Wald")["years",2],
              confint(mod1_commercial, method = "Wald")["years",2],
              confint(mod2_commercial, method = "Wald")["years",2],
              confint(mod1_medicaid, method = "Wald")["years",2],
              confint(mod2_medicaid, method = "Wald")["years",2],
              confint(mod1_uninsured, method = "Wald")["years",2],
              confint(mod2_uninsured, method = "Wald")["years",2])
  )
```

## Table of Results So Far

```{r}
res12
```

## Figure 6. Graph of Results So Far (Model 1: No Covariates)

```{r}
fig_model1 <- res12 %>% filter(model == 1)

fig_model1$subgroup <- fct_reorder(fig_model1$subgroup, fig_model1$estimate)

ggplot(fig_model1, aes(x = estimate, y = subgroup, label = decim(estimate, 2))) +
  geom_vline(xintercept = 0, lty = "dashed", col = "red") +
  geom_errorbarh(aes(xmax = hi.95ci, xmin = low.95ci, height = 0.4)) +
  geom_label() + 
  theme_bw() +
  labs(x = "Estimated Effect on BP < 140/90 (in percentage points)",
       y = "Group",
       title = "No Covariates Model Estimates of Effect Per Year on BP < 140/90",
       subtitle = "in percentage points with 95% confidence intervals",
       caption = "Includes BHP data from practices with 50+ patients, 2012-13 to 2016-17")
```

## Figure 7. Graph of Results So Far (Model 2: Time-Varying Covariates)

```{r}
fig_model2 <- res12 %>% filter(model == 2)

fig_model2$subgroup <- fct_reorder(fig_model2$subgroup, fig_model2$estimate)

ggplot(fig_model2, aes(x = estimate, y = subgroup, label = decim(estimate, 2))) +
  geom_vline(xintercept = 0, lty = "dashed", col = "red") +
  geom_errorbarh(aes(xmax = hi.95ci, xmin = low.95ci, height = 0.4)) +
  geom_label() + 
  theme_bw() +
  labs(x = "Estimated Effect on BP < 140/90 (in percentage points)",
       y = "Group",
       title = "Time-Varying Covs. Model Estimates of Effect Per Year on BP < 140/90",
       subtitle = "in percentage points with 95% confidence intervals",
       caption = "Includes BHP data from practices with 50+ patients, 2012-13 to 2016-17")
```

# Next Steps

Not shown here.

# Session Info

```{r}
devtools::session_info()
```
