---
title: "431 Class 23"
author: "Thomas E. Love"
date: "2017-11-16"
output:
  beamer_presentation:
    theme: "Madrid"
    fonttheme: "structurebold"
    colortheme: "whale"
    fig_caption: false
---

```{r set-options, echo=FALSE, cache=FALSE, message=FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 55)
```

## Today's Agenda

- Discussing Past Activities
    + Quiz 2
    + Assignment 5
    + "After Class 22" Google Form
- Changes in the Tidyverse
- Regression Modeling in the Tidyverse
    + The National Youth Fitness Survey
    
## Today's R Setup and Data Set

```{r setup}
library(magrittr); library(broom); library(tidyverse)

nyfs1 <- read.csv("data/nyfs1.csv") %>% tbl_df

source("Love-boost.R")
```


# Regression, or "You've Got To Draw The Line Somewhere"

## The National Youth Fitness Survey, 2012

See https://thomaselove.github.io/431notes/NYFS-Study.html for details.

```{r}
summary(select(nyfs1, bmi, waist.circ, sex))
```

## Summarizing Associations Graphically (Code)

```{r, eval = FALSE}
cor_lab <- nyfs1 %$% 
  cor(waist.circ, bmi) %>% 
  round(.,2)

ggplot(nyfs1, aes(x = waist.circ, y = bmi)) +
  geom_point(pch = 1) +
  geom_smooth(method = "lm", col = "red") + 
  geom_smooth(method = "loess", col = "blue") + 
  labs(x = "Waist Circumference (cm)", 
       y = "Body Mass Index") +
  annotate("text", x = 110, y = 10, size = 5,
           label = paste("r =", 
                         cor_lab))
```

Plot on next slide...

## Summarizing Associations Graphically

```{r, echo = FALSE}
cor_lab <- nyfs1 %$% 
  cor(waist.circ, bmi) %>% 
  round(.,2)

ggplot(nyfs1, aes(x = waist.circ, y = bmi)) +
  geom_point(pch = 1) +
  geom_smooth(method = "lm", col = "red") + 
  geom_smooth(method = "loess", col = "blue") + 
  labs(x = "Waist Circumference (cm)", 
       y = "Body Mass Index") +
  annotate("text", x = 110, y = 10, size = 5,
           label = paste("r =", 
                         cor_lab))
```

## Summarizing Associations Numerically

```{r}
nyfs1 %$% cor(waist.circ, bmi)
```

### The **Pearson** Correlation, *r*

- Unitless (scale-free) measure of bivariate linear association
- $r_{XY}$ = (slope of Y~X regression line) times SD~X~ / SD~Y~
- -1 $\leq$ *r* $\leq$ +1
    - -1 indicates straight line relationship with negative slope
    - +1 indicates straight line relationship with positive slope
    - 0 indicates no linear association
- r^2^ = key regression summary - % of variation in Y explained by X

## Scatterplot Matrix for some NYFS1 variables (Code)

```{r, eval = FALSE, message = FALSE}
GGally::ggpairs(select(nyfs1, 
                       age.exam, triceps.skinfold, 
                       sex, waist.circ, bmi), 
        title = "Scatterplot Matrix for nyfs1 data")
```

Note that I usually run this with `message = FALSE` in the chunk label. Otherwise, you get some irritating messages in your output.

## Scatterplot Matrix for some NYFS1 variables

```{r, echo = FALSE, message = FALSE}
GGally::ggpairs(select(nyfs1, 
                       age.exam, sex, waist.circ, bmi), 
        title = "Scatterplot Matrix for nyfs1 data")
```

# Model 1: Predicting `bmi` using `waist.circ` alone

## Summarizing the `bmi`-`waist.circ` association

```{r, echo = FALSE}
cor_lab <- nyfs1 %$% 
  cor(waist.circ, bmi) %>% 
  round(.,2)

ggplot(nyfs1, aes(x = waist.circ, y = bmi)) +
  geom_point(pch = 1) +
  geom_smooth(method = "lm", col = "red") + 
  geom_smooth(method = "loess", col = "blue") + 
  labs(x = "Waist Circumference (cm)", 
       y = "Body Mass Index") +
  annotate("text", x = 110, y = 10, size = 5,
           label = paste("r =", 
                         cor_lab))
```

## Linear Model 1: `bmi` ~ f(`waist.circ`)

```{r}
m1 <- nyfs1 %$% lm(bmi ~ waist.circ)

m1
```

```{r, eval = FALSE}
summary(m1) ## see next slide
```

## Summary for Linear Model (rearranged, a bit)

```
Call: lm(formula = bmi ~ waist.circ)

Coefficients: Estimate Std. Error t value Pr(>|t|)    
(Intercept)   -0.06646    0.23292  -0.285    0.775    
waist.circ     0.28893    0.00350  82.548   <2e-16 ***
---
Sig. codes: 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Multiple R-squared:  0.8282
F-statistic:  6814 on 1 and 1414 DF,  p-value: < 2.2e-16

Residuals:     Min      1Q  Median      3Q     Max 
           -4.2343 -1.0941 -0.0744  0.9254 19.0664 
Residual standard error: 1.692 on 1414 degrees of freedom
Adjusted R-squared:  0.828 
```

## 95% Confidence Intervals for `m1` Coefficients

```{r}
m1
confint(m1, level = 0.95)
```

## Tidying the Coefficients (with `broom::tidy`)

Places the coefficient summary into a tibble.

```{r}
tidy(m1) ## from broom package
```


## Model Summaries, at a glance (with `broom::glance`)

Places detailed model summaries into a one-row tibble.

```{r}
glance(m1) ## also from broom
```

## What if we tried log(BMI)?

```{r, echo = FALSE}
cor_lab2 <- nyfs1 %$% 
  cor(waist.circ, log(bmi)) %>% 
  round(.,4)

ggplot(nyfs1, aes(x = waist.circ, y = log(bmi))) +
  geom_point(pch = 1) +
  geom_smooth(method = "lm", col = "red") + 
  geom_smooth(method = "loess", col = "blue") + 
  labs(x = "Waist Circumference (cm)", 
       y = "log of Body Mass Index") +
  annotate("text", x = 100, y = 2.75, size = 5,
           label = paste("Now, r =", cor_lab2))

```


## log(BMI) vs. BMI: The effect of the log

```{r, echo = FALSE}
ggplot(nyfs1, aes(x = bmi, y = log(bmi))) + 
  geom_point(pch = 1) +
  labs(x = "Body-Mass Index", y = "Natural log of BMI")
```

## But what if we tried modeling the log of BMI?

```{r}
m1.log <- nyfs1 %$% lm(log(bmi) ~ waist.circ)

m1.log
```

## `m1.log` vs. `m1` (Coefficients)

```{r}
tidy(m1)
tidy(m1.log)
```

## `m1.log` vs. `m1` (Summaries)

```{r}
glance(m1)
glance(m1.log)
```


## Augmenting the Data with Model `m1` Results (`broom::augment`) run with `warning = FALSE`

```{r, warning = FALSE}
newdata <- augment(m1) ## yet again from broom

head(newdata,3)
```

## Assumption Checking: The Role of Residuals

1. ("Linearity") The residuals from the linear model should show no particular curved relationship when plotted against the fitted values, or when plotted against individual predictors.
2. ("Independence") Particularly if the data describe a series in time or space, we want to see no clear cycles in the residuals when plotted against time/space.
3. ("Homoscedasticity" / "Constant Variance") The residuals from the linear model should display a similar amount of spread across levels of the fitted values. Sometimes we plot the square root of the residuals against the fitted values to assess this.
4. ("Approximately Normal") The residuals from the linear model should follow an approximately Normal distribution. Often, we'll plot the residuals in a normal Q-Q plot to assess this. We'd like to avoid [a] fitting points poorly (exceptionally large positive or negative residuals), [b] points that show enormous influence over the model (removing them causes the coefficients to change substantially)

## Plotting the Model Residuals (squeezed too much)

```{r, fig.height = 4}
par(mfrow = c(2,2))
plot(m1)
par(mfrow = c(1,1))
```

## `plot(m1, which = 1)`    Residuals vs Fitted Values 

```{r, echo = FALSE}
plot(m1, which = 1)
```

## `plot(m1, which = 3)`   Scale-Location Plot

```{r, echo = FALSE}
plot(m1, which = 3)
```

## Subjects labeled 583, 762 and 1156

```{r}
nyfs1 %>% slice(c(583, 762, 1156)) %>%
  select(subject.id, bmi, waist.circ)
```

What about our augmented data?

## Subjects labeled 583, 762 and 1156

```{r, warning = FALSE}
augment(m1) %>%
  slice(c(583, 762, 1156)) %>%
  select(bmi, waist.circ, .fitted, .resid, 
         .std.resid, .cooksd) %>%
  round(., 2)
```

## Which point has the largest residual in the data?

```{r, echo = FALSE}
ggplot(nyfs1, aes(x = waist.circ, y = bmi)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", col = "red") + 
  labs(x = "Waist Circumference (cm)", 
       y = "Body Mass Index",
       title = "Which point has the largest residual?") 
```

## Row 1156 (ID 73202) has the largest residual

```{r, echo = FALSE}
temp <- nyfs1
temp$highlight <- ifelse(temp$subject.id == 73202, "1156 (id 73202)", "all others")

ggplot(temp, aes(x = waist.circ, y = bmi, col = highlight)) +
  geom_point(size = 3) +
  guides(col = FALSE) +
  geom_smooth(method = "lm", col = "red") +
  theme_bw() +
  labs(x = "Waist Circumference",
       y = "BMI", 
       title = "In red, subject 73202 (row 1156) - largest residual")
```


## `plot(m1, which = 2)`

```{r, echo = FALSE}
plot(m1, which = 2)
```

## Standardized Residuals just rescale Raw Residuals

```{r, echo = FALSE}
par(mfrow = c(2,1))
boxplot(newdata$.resid, xlab = ".resid = Model m1 Raw Residuals", main = "Boxplot of m1 .resid", horizontal = TRUE)
boxplot(newdata$.std.resid, xlab = ".std.resid = Model m1 Standardized Residuals", main = "Boxplot of m1 .std.resid", horizontal = TRUE)
par(mfrow=c(1,1))
```

## `.std.resid` from Model `m1` (Notes, Section 39.2)

```{r, echo = FALSE, message = FALSE}
p1 <- ggplot(newdata, aes(x = .std.resid)) +
    geom_histogram(aes(y = ..density..),
                   bins = fd_bins(newdata$.std.resid), 
                   fill = '#440154', col = '#FDE725') +
    stat_function(fun = dnorm,
                  args = list(mean = mean(newdata$.std.resid), 
                              sd = sd(newdata$.std.resid)), 
                  lwd = 1.5, col = '#1FA187') +
    labs(title = "Histogram", x = "Standardized Residual", y = "") +
    theme_bw() 

p2 <- ggplot(newdata, aes(x = 1, y = .std.resid)) + 
    geom_boxplot(fill = '#FDE725', notch = TRUE, 
                 col = '#440154', outlier.color = '#440154') +
    labs(title = "Boxplot", x = "", y = "") +
    theme_bw() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank())  

p3 <- ggplot(newdata, aes(sample = .std.resid)) + 
    geom_qq(geom = "point", col = '#440154', size = 2) + 
    geom_abline(slope = qq_slope(newdata$.std.resid),
                intercept = qq_int(newdata$.std.resid),
                col = '#1FA187', size = 1.25) +
    labs(title = "Normal QQ", x = "", y = "") +
    theme_bw()

p <- cowplot::plot_grid(p1, p2, p3, align = "h", nrow = 1,
                   rel_widths = c(3, 1, 2))

title <- cowplot::ggdraw() + 
    cowplot::draw_label("Standardized Residuals from Model m1", 
               fontface = "bold")

cowplot::plot_grid(title, p, ncol = 1, rel_heights=c(0.1, 1))
```


## Outliers, Leverage, Influence

We will also examine the **leverage** of individual points (`.hat` in the augment output) where large values indicate unusual combinations of predictors, especially when we have more than one predictor.

```{r, warning = FALSE}
augment(m1) %>%
  slice(c(583, 762, 1156)) %>%
  select(bmi, waist.circ, .hat, 
         .std.resid, .cooksd) 
```

## `plot(m1, which = 5)`   Residuals vs. Leverage Plot

```{r, echo = FALSE}
plot(m1, which = 5)
```

## `plot(m1, which = 4)`   Cook's Distance Plot

```{r, echo = FALSE}
plot(m1, which = 4)
```

## Why does 583 have the largest Cook's distance?

```{r}
summary(select(nyfs1, waist.circ, bmi))
```

## Why does 583 have the largest Cook's distance?

```{r}
nyfs1 %>% slice(583) %>% 
  select(subject.id, waist.circ, bmi)
```

## Why does 583 have the largest Cook's distance?

```{r, echo = FALSE}
temp <- nyfs1
temp$highlight <- ifelse(temp$subject.id == 72560, "583 (id 72560)", "all others")

ggplot(temp, aes(x = waist.circ, y = bmi, col = highlight)) +
  geom_point(size = 3) +
  guides(col = FALSE) +
  geom_smooth(method = "lm", col = "red") +
  theme_bw() +
  labs(x = "Waist Circumference",
       y = "BMI", 
       title = "In red, subject 72560 (row 583) - highest influence (Cook's d)")
```

## Model `m1`, without point 583

```{r}
m1.no583 <- lm(bmi ~ waist.circ, data = slice(nyfs1, -583))

m1.no583
```

## `m1` with and without 583, coefficients

```{r}
tidy(m1) 
tidy(m1.no583) 
```


## `m1` with and without 583, summaries

```{r}
glance(m1)
glance(m1.no583)
```

# Model 2: Predicting `bmi` using `waist.circ` as well as `age.exam` and `sex`

## Scatterplot Matrix for some NYFS1 variables

```{r, echo = FALSE, message = FALSE}
GGally::ggpairs(select(nyfs1, 
                       age.exam, sex, waist.circ, bmi), 
        title = "Scatterplot Matrix for nyfs1 data")
```

## Building Model 2

```{r}
m2 <- nyfs1 %$%
  lm(bmi ~ waist.circ + age.exam + sex)

m2
```

## Summary of Model `m2` (rearranged a little)

```
Call: lm(formula = bmi ~ waist.circ + age.exam + sex)

Multiple R-squared:  0.8635,	Adjusted R-squared:  0.8633 
F-statistic:  2979 on 3 and 1412 DF,  p-value: < 2.2e-16

Coefficients: Estimate Std. Error t value Pr(>|t|)    
(Intercept)  -1.445179   0.222018  -6.509 1.05e-10 ***
waist.circ    0.348370   0.004421  78.801  < 2e-16 ***
age.exam     -0.293247   0.015440 -18.992  < 2e-16 ***
sexMale       0.188094   0.080208   2.345   0.0192 *  
---
Sig. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residuals:     Min      1Q  Median      3Q     Max 
           -4.0232 -0.8879 -0.0802  0.7836 20.3659 
Residual standard error: 1.509 on 1412 degrees of freedom
```

## Summary of `m1` (for reference, rearranged)

```
Call: lm(formula = bmi ~ waist.circ)

Multiple R-squared:  0.8282,	Adjusted R-squared:  0.828 
F-statistic:  6814 on 1 and 1414 DF,  p-value: < 2.2e-16

Coefficients: Estimate Std. Error t value Pr(>|t|)    
(Intercept)   -0.06646    0.23292  -0.285    0.775    
waist.circ     0.28893    0.00350  82.548   <2e-16 ***
---
Sig. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residuals:     Min      1Q  Median      3Q     Max 
           -4.2343 -1.0941 -0.0744  0.9254 19.0664 
Residual standard error: 1.692 on 1414 degrees of freedom
```

## 95% Confidence Intervals for `m2` Coefficients

```{r}
confint(m2, level = 0.95)
```

## Tidying the Coefficients (with `broom::tidy`)

Places the coefficient summary into a tibble.

```{r}
tidy(m2) ## from broom package
```


## Model Summaries, at a glance (with `broom::glance`)

```{r}
glance(m2) ## also from broom
glance(m1) ## for comparison
```

## Augmenting the Data with Model Results (`broom::augment`) run with `warning = FALSE`

```{r, warning = FALSE}
newdat2 <- augment(m2) ## again from broom
head(newdat2,3)
```

## Residuals vs. Fitted Values `plot(m2, which = 1)`

```{r, echo = FALSE}
plot(m2, which = 1)
```

## Scale-Location Plot `plot(m2, which = 3)`

```{r, echo = FALSE}
plot(m2, which = 3)
```

## Standardized Residuals `plot(m2, which = 2)`

```{r, echo = FALSE}
plot(m2, which = 2)
```

## Residuals, Leverage, Influence `plot(m2, which = 5)`

```{r, echo = FALSE}
plot(m2, which = 5)
```

## Cook's Distance Index Plot `plot(m2, which = 4)`

```{r, echo = FALSE}
plot(m2, which = 4)
```

## Coming Soon (and see Notes, Sections 37-end)

- What happens when a linear model doesn't fit so well?
    + Should we be modeling a transformed outcome?
    + Box-Cox approach to transformation decisions
- Incorporating multi-categorical predictors
- Model Validation: Making predictions in new data
    + Simplest Smart Approach: Split data into Training, Test samples
- Variable Selection: Stepwise and better approaches
- Standardizing Regression Coefficients
- Dealing with Missingness Sensibly
